---
title: "Learning Analytics MOOC Engagement Analysis"
subtitle: "MAS8600/MAS8505 Coursework"
author: "Pranit Chatterjee"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)


library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(knitr)
library(kableExtra)
```

---

# Executive Summary

This analysis examines learner engagement patterns across 7 runs of a MOOC using data from 37,296 enrollment records. Using CRISP-DM methodology, we identify key engagement drivers and completion barriers. Early findings indicate low completion rates (~5.8%), with significant demographic data gaps (>88% missing). Most unenrolled learners disengage within 72 days of enrollment.

---

# 1. Business Understanding

## 1.1 Business Objective

The primary objective is to understand MOOC learner engagement and identify patterns in:

- **Course completion rates** by learner cohort  
- **Unenrollment timing** and duration patterns  
- **Certificate purchase behavior** (indicator of serious engagement)  
- **Demographic factors** influencing engagement  

## 1.2 Success Criteria

- Identify engagement segments (completers, active, unenrolled).  
- Quantify completion rates and conversion to certificates.  
- Provide actionable insights for retention strategies.  

## 1.3 Data Overview

- **Total enrollments:** 37,296 learners across 7 course runs  
- **Time period:** 2016–2018  
- **Key variables:** Enrollment, unenrollment, course completion, demographics  

---

# 2. Data Understanding

## 2.1 Data Quality Assessment

### Data Loading

```{r}

project_root <- normalizePath("..")
# Load the 7 CSVs directly (adjust paths if needed)
enrollment_1 <- read_csv(file.path(project_root, "data/enrollment_1.csv")) %>% mutate(course_run = 1)
enrollment_2 <- read_csv(file.path(project_root, "data/enrollment_2.csv")) %>% mutate(course_run = 2)
enrollment_3 <- read_csv(file.path(project_root, "data/enrollment_3.csv")) %>% mutate(course_run = 3)
enrollment_4 <- read_csv(file.path(project_root, "data/enrollment_4.csv")) %>% mutate(course_run = 4)
enrollment_5 <- read_csv(file.path(project_root, "data/enrollment_5.csv")) %>% mutate(course_run = 5)
enrollment_6 <- read_csv(file.path(project_root, "data/enrollment_6.csv")) %>% mutate(course_run = 6)
enrollment_7 <- read_csv(file.path(project_root, "data/enrollment_7.csv")) %>% mutate(course_run = 7)

all_enrollments <- dplyr::bind_rows(
  enrollment_1,
  enrollment_2,
  enrollment_3,
  enrollment_4,
  enrollment_5,
  enrollment_6,
  enrollment_7
)

```

### Feature Engineering

```{r feature-engineering}
all_enrollments <- all_enrollments %>%
  mutate(
    enrolled_date = as.Date(enrolled_at),
    unenrolled_date = as.Date(unenrolled_at),
    enrollment_duration = as.numeric(unenrolled_date - enrolled_date),
    fully_participated = !is.na(fully_participated_at),
    purchased_statement = !is.na(purchased_statement_at),
    unenrolled = !is.na(unenrolled_at),
    engagement_status = case_when(
      fully_participated ~ "Completed",
      unenrolled ~ "Unenrolled",
      TRUE ~ "Still Active / Unknown"
    )
  )
```
### Missing Values Summary

```{r missing-values, results='asis'}
missing_summary <- all_enrollments %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Column", values_to = "Missing_Count") %>%
  mutate(
    Missing_Pct = round(Missing_Count / nrow(all_enrollments) * 100, 1),
    Data_Quality = case_when(
      Missing_Pct == 0 ~ "Complete",
      Missing_Pct < 10 ~ "Good",
      Missing_Pct < 50 ~ "Fair",
      TRUE ~ "Poor"
    )
  ) %>%
  arrange(desc(Missing_Pct))

kable(missing_summary, 
      caption = "Data Quality Assessment: Missing Values",
      align = "lrrl") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE)
```

**Interpretation:** Temporal variables (enrollment, unenrollment, completion dates) are 100% complete. Demographic variables show >88% missing, indicating limited self-reported data from learners.

### "Unknown" Values in Categorical Fields

```{r unknown-values, results='asis'}
unknown_summary <- tibble(
  Field = c("Gender", "Age Range", "Education Level", "Employment Status"),
  Unknown_Count = c(
    sum(all_enrollments$gender == "Unknown", na.rm = TRUE),
    sum(all_enrollments$age_range == "Unknown", na.rm = TRUE),
    sum(all_enrollments$highest_education_level == "Unknown", na.rm = TRUE),
    sum(all_enrollments$employment_status == "Unknown", na.rm = TRUE)
  )
) %>%
  mutate(
    Pct_Total = round(Unknown_Count / nrow(all_enrollments) * 100, 1)
  )

kable(unknown_summary,
      caption = "Unknown/Not Specified Values in Demographic Fields",
      align = "lrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## 2.2 Enrollment Timeline

```{r date-range}
# ensure enrolled_date is Date
all_enrollments$enrolled_date <- as.Date(all_enrollments$enrolled_at)

min_date <- min(all_enrollments$enrolled_date, na.rm = TRUE)
max_date <- max(all_enrollments$enrolled_date, na.rm = TRUE)

cat("**Enrollment Date Range:**\n\n")
cat("- Earliest enrollment:", format(min_date, "%B %d, %Y"), "\n")
cat("- Latest enrollment:", format(max_date, "%B %d, %Y"), "\n")
cat("- Time span:", as.numeric(max_date - min_date), "days (~", 
    round(as.numeric(max_date - min_date) / 365, 1), "years)\n")
```

## 2.3 Key Data Characteristics

```{r data-characteristics, results='asis'}
char_summary <- tibble(
  Metric = c(
    "Total Enrollments",
    "Unique Course Runs",
    "Date Range (days)",
    "Complete Demographics",
    "With Gender",
    "With Age Range",
    "With Education Level",
    "With Employment Status"
  ),
  Value = c(
    format(nrow(all_enrollments), big.mark = ","),
    n_distinct(all_enrollments$course_run),
    as.numeric(max_date - min_date),
    sum(all_enrollments$has_demographics),
    sum(!is.na(all_enrollments$gender) & all_enrollments$gender != "Unknown"),
    sum(!is.na(all_enrollments$age_range) & all_enrollments$age_range != "Unknown"),
    sum(!is.na(all_enrollments$highest_education_level) & all_enrollments$highest_education_level != "Unknown"),
    sum(!is.na(all_enrollments$employment_status) & all_enrollments$employment_status != "Unknown")
  )
)

kable(char_summary,
      caption = "Dataset Characteristics",
      align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

---

# 3. Data Preparation

## 3.1 Data Cleaning Steps

The following transformations were applied:

1. **Date Conversion:** All date-time strings converted to proper Date objects (`enrolled_date`, `unenrolled_date`, `fully_participated_date`, `purchased_statement_date`).

2. **Standardization:** "Unknown" values in demographic fields converted to `NA` for proper handling.

3. **Course Run Coding:** Each enrollment tagged with its course run (1–7) for comparative analysis.

## 3.2 Feature Engineering

New derived variables created for engagement analysis:

| Feature | Definition | Type |
|---------|-----------|------|
| `fully_participated` | TRUE if learner reached course completion milestone | Binary |
| `purchased_statement` | TRUE if learner purchased certificate/statement | Binary |
| `unenrolled` | TRUE if learner unenrolled before course end | Binary |
| `enrollment_duration` | Days between enrollment and unenrollment | Numeric |
| `engagement_status` | Categorical: "Completed", "Unenrolled", "Still Active / Unknown" | Factor |
| `has_demographics` | TRUE if learner provided complete demographic data | Binary |

---

# 4. Exploratory Data Analysis

## 4.1 Engagement Status Distribution

### Summary Statistics

```{r engagement-summary, results='asis'}
all_enrollments <- all_enrollments %>%
  mutate(
    fully_participated = fully_participated %in% c(TRUE, 1, "TRUE"),
    unenrolled = unenrolled %in% c(TRUE, 1, "TRUE"),
    engagement_status = case_when(
      fully_participated ~ "Completed",
      unenrolled ~ "Unenrolled",
      TRUE ~ "Still Active / Unknown"
    )
  )

engagement_dist <- all_enrollments %>%
  count(engagement_status) %>%
  mutate(
    Percentage = round(n / sum(n) * 100, 2),
    Cumulative_Pct = round(cumsum(n) / sum(n) * 100, 2)
  ) %>%
  rename("Engagement Status" = engagement_status, "Count" = n)

kable(engagement_dist,
      caption = "Learner Engagement Status",
      align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

### Visualization: Engagement Status

```{r engagement-plot, fig.cap="Distribution of learners by engagement status"}
all_enrollments %>%
  count(engagement_status) %>%
  mutate(engagement_status = factor(engagement_status, 
                                     levels = c("Completed", "Unenrolled", "Still Active / Unknown"))) %>%
  ggplot(aes(x = reorder(engagement_status, -n), y = n, fill = engagement_status)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "\n(", round(n/sum(n)*100, 1), "%)")),
            vjust = -0.5, size = 4) +
  labs(
    title = "Learner Engagement Status Distribution",
    x = "Engagement Status",
    y = "Number of Learners",
    caption = "Total enrollments: 37,296"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.2 Completion Analysis

### Completion Indicators

```{r completion-summary, results='asis'}
completion_stats <- tibble(
  Indicator = c("Fully Participated", "Purchased Statement", "Unenrolled"),
  Count = c(
    sum(all_enrollments$fully_participated),
    sum(all_enrollments$purchased_statement),
    sum(all_enrollments$unenrolled)
  )
) %>%
  mutate(
    Percentage = round(Count / nrow(all_enrollments) * 100, 2)
  )

kable(completion_stats,
      caption = "Completion and Engagement Indicators",
      align = "lrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Key Insights:**

- Only **5.78% (2,154 learners)** fully participated in the course.
- **0.77% (289 learners)** purchased a certificate/statement, indicating very limited willingness to pay.
- **11.24% (4,191 learners)** unenrolled before course end.
- **82.21% (30,762 learners)** remain in "Still Active / Unknown" status, likely representing informal or discontinued engagement.

### Visualization: Completion Funnel

```{r completion-funnel, fig.cap="Completion funnel showing progression from enrollment to certificate purchase"}
funnel_data <- tibble(
  Stage = c("Enrolled", "Fully Participated", "Purchased Statement"),
  Count = c(
    nrow(all_enrollments),
    sum(all_enrollments$fully_participated),
    sum(all_enrollments$purchased_statement)
  )
) %>%
  mutate(
    Percentage = round(Count / nrow(all_enrollments) * 100, 2),
    Stage = factor(Stage, levels = c("Enrolled", "Fully Participated", "Purchased Statement"))
  )

ggplot(funnel_data, aes(x = Stage, y = Count, fill = Stage)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(Count, "\n(", Percentage, "%)")),
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Course Completion Funnel",
    subtitle = "Drop-off at each progression stage",
    x = "Progression Stage",
    y = "Number of Learners"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.3 Unenrollment Duration Analysis

### Duration Statistics

```{r duration-stats, results='asis'}
duration_summary <- all_enrollments %>%
  filter(!is.na(enrollment_duration)) %>%
  summarise(
    Metric = c("Mean", "Median", "Std. Dev", "Min", "Max", "Q1", "Q3"),
    Value_Days = c(
      round(mean(enrollment_duration), 1),
      round(median(enrollment_duration), 1),
      round(sd(enrollment_duration), 1),
      round(min(enrollment_duration), 1),
      round(max(enrollment_duration), 1),
      round(quantile(enrollment_duration, 0.25), 1),
      round(quantile(enrollment_duration, 0.75), 1)
    )
  )

kable(duration_summary,
      caption = "Unenrollment Duration Statistics (Days)",
      align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Interpretation:** Unenrolled learners persist for an average of **124 days** (median: **72 days**), with high variability (SD: 145 days). This suggests a critical drop-off window around 2–3 months post-enrollment.

### Visualization: Duration Distribution

```{r duration-histogram, fig.cap="Distribution of days from enrollment to unenrollment"}
all_enrollments %>%
  filter(!is.na(enrollment_duration) & enrollment_duration > 0) %>%
  ggplot(aes(x = enrollment_duration)) +
  geom_histogram(binwidth = 30, fill = "#2E86AB", alpha = 0.8, color = "white") +
  geom_vline(aes(xintercept = median(enrollment_duration, na.rm = TRUE)),
             color = "#A23B72", linetype = "dashed", size = 1, label = "Median") +
  geom_vline(aes(xintercept = mean(enrollment_duration, na.rm = TRUE)),
             color = "#F18F01", linetype = "dashed", size = 1, label = "Mean") +
  labs(
    title = "Time to Unenrollment Distribution",
    x = "Days from Enrollment to Unenrollment",
    y = "Number of Learners",
    caption = "Excludes 0-day unenrollments (same-day drop)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  ) +
  scale_x_continuous(breaks = seq(0, 900, 150))
```

### Visualization: Unenrollment by Duration Quartile

```{r duration-quartiles, fig.cap="Breakdown of unenrollments by duration quartiles"}
all_enrollments %>%
  filter(!is.na(enrollment_duration)) %>%
  mutate(
    Duration_Quartile = case_when(
      enrollment_duration <= 7 ~ "≤ 1 week (0-7 days)",
      enrollment_duration <= 30 ~ "1-4 weeks (8-30 days)",
      enrollment_duration <= 72 ~ "1-10 weeks (31-72 days)",
      enrollment_duration <= 180 ~ "2-6 months (73-180 days)",
      TRUE ~ "> 6 months (>180 days)"
    ),
    Duration_Quartile = factor(Duration_Quartile,
                               levels = c("≤ 1 week (0-7 days)",
                                          "1-4 weeks (8-30 days)",
                                          "1-10 weeks (31-72 days)",
                                          "2-6 months (73-180 days)",
                                          "> 6 months (>180 days)"))
  ) %>%
  count(Duration_Quartile) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  ggplot(aes(x = Duration_Quartile, y = n, fill = Duration_Quartile)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "\n(", Percentage, "%)")),
            vjust = -0.5, size = 3.5) +
  labs(
    title = "Unenrollment by Duration Quartile",
    x = "Time to Unenrollment",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.4 Engagement by Course Run

### Completion Rates Across Runs

```{r run-comparison, results='asis'}
run_stats <- all_enrollments %>%
  group_by(course_run) %>%
  summarise(
    Enrollments = n(),
    Completed = sum(fully_participated),
    Completion_Rate = round(sum(fully_participated) / n() * 100, 2),
    Unenrolled = sum(unenrolled),
    Unenrollment_Rate = round(sum(unenrolled) / n() * 100, 2),
    Purchased = sum(purchased_statement),
    Purchase_Rate = round(sum(purchased_statement) / n() * 100, 3),
    Avg_Duration = round(mean(enrollment_duration, na.rm = TRUE), 1)
  ) %>%
  rename(
    "Course Run" = course_run,
    "Enroll #" = Enrollments,
    "Complete #" = Completed,
    "Complete %" = Completion_Rate,
    "Unenroll #" = Unenrolled,
    "Unenroll %" = Unenrollment_Rate,
    "Purchase #" = Purchased,
    "Purchase %" = Purchase_Rate,
    "Avg Days" = Avg_Duration
  )

kable(run_stats,
      caption = "Engagement Metrics by Course Run",
      align = "lrrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

### Visualization: Completion Rate by Run

```{r run-completion-plot, fig.cap="Completion rates vary across course runs"}
all_enrollments %>%
  group_by(course_run) %>%
  summarise(
    Completion_Rate = sum(fully_participated) / n() * 100,
    Enrollments = n()
  ) %>%
  ggplot(aes(x = course_run, y = Completion_Rate, fill = course_run)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Completion_Rate, 2), "%")),
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Course Completion Rate by Run",
    x = "Course Run",
    y = "Completion Rate (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(size = 11)
  ) +
  scale_y_continuous(limits = c(0, max(all_enrollments %>%
    group_by(course_run) %>%
    summarise(Completion_Rate = sum(fully_participated) / n() * 100) %>%
    pull(Completion_Rate)) * 1.15))
```

---

# 5. Key Findings & Insights

## 5.1 Engagement Patterns

1. **Very Low Completion:** Only 5.78% of enrollments result in full course participation, indicating a significant engagement challenge.

2. **Low Monetization:** Certificate/statement purchase rate is extremely low (0.77%), suggesting limited perceived value or willingness to pay among learners.

3. **High Unenrollment:** 11.24% of learners actively unenroll, with a median time of **72 days**—indicating most drop-off occurs in the first 2–3 months.

4. **Large "Unknown" Cohort:** 82.21% of enrollments fall into "Still Active / Unknown," likely representing learners who stop engaging informally without formal unenrollment.

## 5.2 Critical Drop-Off Window

The unenrollment duration analysis reveals:

- **41% of unenrolled learners** drop out within the first week.  
- **Median persistence:** 72 days (approximately 10 weeks).  
- **Max persistence:** 889 days (~2.4 years), indicating some long-term lurkers.  

This suggests an **early intervention window (weeks 1–10)** is critical for retention.

## 5.3 Data Quality Implications

High missing demographic data (>88%) limits demographic segmentation. Analysis cannot confidently compare engagement by age, education, or employment without imputation or assumptions.

---

# 6. Conclusions & Recommendations

## 6.1 Conclusions

1. **MOOC completion is rare** (~5.8%), and conversion to paid certificates is minimal (~0.8%).  
2. **Most disengagement is informal**—learners don't formally unenroll; they simply stop participating.  
3. **Early weeks are critical**—42% of unenrollments occur within 1 week; the median is ~10 weeks.  
4. **Demographic factors are unmeasured**—missing data prevents deeper segmentation analysis.

## 6.2 Recommendations for Next Cycle

1. **Improve learner onboarding** in weeks 1–2 to reduce immediate drop-off.  
2. **Implement engagement checkpoints** at week 2, 4, and 8 to re-activate at-risk learners.  
3. **Collect demographic data** through mandatory profile completion to enable segmentation analysis.  
4. **Investigate "Still Active" cohort** to distinguish true lurkers from inactive accounts.  
5. **Pilot retention interventions** (e.g., peer support, milestone rewards) for next course run.

---
