---
title: "Learning Analytics Report"
author: "Pranit Chatterjee"
date: "2026-01-14"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 300
)


library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(knitr)
library(kableExtra)
library(multcomp)
library(car)
```


# Executive Summary

This analysis examines learner engagement patterns across 7 runs of a MOOC using data from 37,296 enrollment records. Using CRISP-DM methodology, we identify key engagement drivers and completion barriers. Early findings indicate low completion rates (~5.8%), with significant demographic data gaps (>88% missing). Most unenrolled learners disengage within 72 days of enrollment.


# 1. Business Understanding

## 1.1 Business Objective

The primary objective is to understand MOOC learner engagement and identify patterns in:

- **Course completion rates** by learner cohort  
- **Unenrollment timing** and duration patterns  
- **Certificate purchase behavior** (indicator of serious engagement)  
- **Demographic factors** influencing engagement  

## 1.2 Success Criteria

- Identify engagement segments (completers, active, unenrolled).  
- Quantify completion rates and conversion to certificates.  
- Provide actionable insights for retention strategies.  

## 1.3 Data Overview

- **Total enrollments:** 37,296 learners across 7 course runs  
- **Time period:** 2016–2018  
- **Key variables:** Enrollment, unenrollment, course completion, demographics  

# 2. Data Understanding (Cycle - 1)

## 2.1 Data Quality Assessment

### Data Loading

```{r}

project_root <- normalizePath("..")
# Load the 7 CSVs directly (adjust paths if needed)
enrollment_1 <- read_csv(file.path(project_root, "data/enrollment_1.csv")) %>% mutate(course_run = 1)
enrollment_2 <- read_csv(file.path(project_root, "data/enrollment_2.csv")) %>% mutate(course_run = 2)
enrollment_3 <- read_csv(file.path(project_root, "data/enrollment_3.csv")) %>% mutate(course_run = 3)
enrollment_4 <- read_csv(file.path(project_root, "data/enrollment_4.csv")) %>% mutate(course_run = 4)
enrollment_5 <- read_csv(file.path(project_root, "data/enrollment_5.csv")) %>% mutate(course_run = 5)
enrollment_6 <- read_csv(file.path(project_root, "data/enrollment_6.csv")) %>% mutate(course_run = 6)
enrollment_7 <- read_csv(file.path(project_root, "data/enrollment_7.csv")) %>% mutate(course_run = 7)

all_enrollments <- dplyr::bind_rows(
  enrollment_1,
  enrollment_2,
  enrollment_3,
  enrollment_4,
  enrollment_5,
  enrollment_6,
  enrollment_7
)

```

### Feature Engineering

```{r feature-engineering}
all_enrollments <- all_enrollments %>%
  mutate(
    enrolled_date = as.Date(enrolled_at),
    unenrolled_date = as.Date(unenrolled_at),
    enrollment_duration = as.numeric(unenrolled_date - enrolled_date),
    fully_participated = !is.na(fully_participated_at),
    purchased_statement = !is.na(purchased_statement_at),
    unenrolled = !is.na(unenrolled_at),
    engagement_status = case_when(
      fully_participated ~ "Completed",
      unenrolled ~ "Unenrolled",
      TRUE ~ "Still Active / Unknown"
    ),
    has_demographics = !(gender == "Unknown" |
                         age_range == "Unknown" |
                         highest_education_level == "Unknown" |
                         employment_status == "Unknown")
  )

cat("Data loaded and prepared.\n")
cat("Total enrollments:", nrow(all_enrollments), "\n")
cat("Course runs:", n_distinct(all_enrollments$course_run), "\n")
```
### Missing Values Summary

```{r missing-values, results='asis'}
missing_summary <- all_enrollments %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Column", values_to = "Missing_Count") %>%
  mutate(
    Missing_Pct = round(Missing_Count / nrow(all_enrollments) * 100, 1),
    Data_Quality = case_when(
      Missing_Pct == 0 ~ "Complete",
      Missing_Pct < 10 ~ "Good",
      Missing_Pct < 50 ~ "Fair",
      TRUE ~ "Poor"
    )
  ) %>%
  arrange(desc(Missing_Pct))

kable(missing_summary, 
      caption = "Data Quality Assessment: Missing Values",
      align = "lrrl") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), 
                full_width = FALSE)
```

**Interpretation:** Temporal variables (enrollment, unenrollment, completion dates) are 100% complete. Demographic variables show >88% missing, indicating limited self-reported data from learners.

### "Unknown" Values in Categorical Fields

```{r unknown-values, results='asis'}
unknown_summary <- tibble(
  Field = c("Gender", "Age Range", "Education Level", "Employment Status"),
  Unknown_Count = c(
    sum(all_enrollments$gender == "Unknown", na.rm = TRUE),
    sum(all_enrollments$age_range == "Unknown", na.rm = TRUE),
    sum(all_enrollments$highest_education_level == "Unknown", na.rm = TRUE),
    sum(all_enrollments$employment_status == "Unknown", na.rm = TRUE)
  )
) %>%
  mutate(
    Pct_Total = round(Unknown_Count / nrow(all_enrollments) * 100, 1)
  )

kable(unknown_summary,
      caption = "Unknown/Not Specified Values in Demographic Fields",
      align = "lrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## 2.2 Enrollment Timeline

```{r date-range}
# ensure enrolled_date is Date
all_enrollments$enrolled_date <- as.Date(all_enrollments$enrolled_at)

min_date <- min(all_enrollments$enrolled_date, na.rm = TRUE)
max_date <- max(all_enrollments$enrolled_date, na.rm = TRUE)

cat("**Enrollment Date Range:**\n\n")
cat("- Earliest enrollment:", format(min_date, "%B %d, %Y"), "\n")
cat("- Latest enrollment:", format(max_date, "%B %d, %Y"), "\n")
cat("- Time span:", as.numeric(max_date - min_date), "days (~", 
    round(as.numeric(max_date - min_date) / 365, 1), "years)\n")
```

## 2.3 Key Data Characteristics

```{r data-characteristics, results='asis'}
char_summary <- tibble(
  Metric = c(
    "Total Enrollments",
    "Unique Course Runs",
    "Date Range (days)",
    "Complete Demographics",
    "With Gender",
    "With Age Range",
    "With Education Level",
    "With Employment Status"
  ),
  Value = c(
    format(nrow(all_enrollments), big.mark = ","),
    n_distinct(all_enrollments$course_run),
    as.numeric(max_date - min_date),
    sum(all_enrollments$has_demographics),
    sum(!is.na(all_enrollments$gender) & all_enrollments$gender != "Unknown"),
    sum(!is.na(all_enrollments$age_range) & all_enrollments$age_range != "Unknown"),
    sum(!is.na(all_enrollments$highest_education_level) & all_enrollments$highest_education_level != "Unknown"),
    sum(!is.na(all_enrollments$employment_status) & all_enrollments$employment_status != "Unknown")
  )
)

kable(char_summary,
      caption = "Dataset Characteristics",
      align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```


# 3. Data Preparation

## 3.1 Data Cleaning Steps

The following transformations were applied:

1. **Date Conversion:** All date-time strings converted to proper Date objects (`enrolled_date`, `unenrolled_date`, `fully_participated_date`, `purchased_statement_date`).

2. **Standardization:** "Unknown" values in demographic fields converted to `NA` for proper handling.

3. **Course Run Coding:** Each enrollment tagged with its course run (1–7) for comparative analysis.

## 3.2 Feature Engineering

New derived variables created for engagement analysis:

| Feature | Definition | Type |
|---------|-----------|------|
| `fully_participated` | TRUE if learner reached course completion milestone | Binary |
| `purchased_statement` | TRUE if learner purchased certificate/statement | Binary |
| `unenrolled` | TRUE if learner unenrolled before course end | Binary |
| `enrollment_duration` | Days between enrollment and unenrollment | Numeric |
| `engagement_status` | Categorical: "Completed", "Unenrolled", "Still Active / Unknown" | Factor |
| `has_demographics` | TRUE if learner provided complete demographic data | Binary |


# 4. Exploratory Data Analysis

## 4.1 Engagement Status Distribution

### Summary Statistics

```{r engagement-summary, results='asis'}
all_enrollments <- all_enrollments %>%
  mutate(
    fully_participated = fully_participated %in% c(TRUE, 1, "TRUE"),
    unenrolled = unenrolled %in% c(TRUE, 1, "TRUE"),
    engagement_status = case_when(
      fully_participated ~ "Completed",
      unenrolled ~ "Unenrolled",
      TRUE ~ "Still Active / Unknown"
    )
  )

engagement_dist <- all_enrollments %>%
  count(engagement_status) %>%
  mutate(
    Percentage = round(n / sum(n) * 100, 2),
    Cumulative_Pct = round(cumsum(n) / sum(n) * 100, 2)
  ) %>%
  rename("Engagement Status" = engagement_status, "Count" = n)

kable(engagement_dist,
      caption = "Learner Engagement Status",
      align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

### Visualization: Engagement Status

```{r engagement-plot, fig.cap="Distribution of learners by engagement status"}
all_enrollments %>%
  count(engagement_status) %>%
  mutate(engagement_status = factor(engagement_status, 
                                     levels = c("Completed", "Unenrolled", "Still Active / Unknown"))) %>%
  ggplot(aes(x = reorder(engagement_status, -n), y = n, fill = engagement_status)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "\n(", round(n/sum(n)*100, 1), "%)")),
            vjust = -0.5, size = 4) +
  labs(
    title = "Learner Engagement Status Distribution",
    x = "Engagement Status",
    y = "Number of Learners",
    caption = "Total enrollments: 37,296"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.2 Completion Analysis

### Completion Indicators

```{r completion-summary, results='asis'}
completion_stats <- tibble(
  Indicator = c("Fully Participated", "Purchased Statement", "Unenrolled"),
  Count = c(
    sum(all_enrollments$fully_participated),
    sum(all_enrollments$purchased_statement),
    sum(all_enrollments$unenrolled)
  )
) %>%
  mutate(
    Percentage = round(Count / nrow(all_enrollments) * 100, 2)
  )

kable(completion_stats,
      caption = "Completion and Engagement Indicators",
      align = "lrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Key Insights:**

- Only **5.78% (2,154 learners)** fully participated in the course.
- **0.77% (289 learners)** purchased a certificate/statement, indicating very limited willingness to pay.
- **11.24% (4,191 learners)** unenrolled before course end.
- **82.21% (30,762 learners)** remain in "Still Active / Unknown" status, likely representing informal or discontinued engagement.

### Visualization: Completion Funnel

```{r completion-funnel, fig.cap="Completion funnel showing progression from enrollment to certificate purchase"}
funnel_data <- tibble(
  Stage = c("Enrolled", "Fully Participated", "Purchased Statement"),
  Count = c(
    nrow(all_enrollments),
    sum(all_enrollments$fully_participated),
    sum(all_enrollments$purchased_statement)
  )
) %>%
  mutate(
    Percentage = round(Count / nrow(all_enrollments) * 100, 2),
    Stage = factor(Stage, levels = c("Enrolled", "Fully Participated", "Purchased Statement"))
  )

ggplot(funnel_data, aes(x = Stage, y = Count, fill = Stage)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(Count, "\n(", Percentage, "%)")),
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Course Completion Funnel",
    subtitle = "Drop-off at each progression stage",
    x = "Progression Stage",
    y = "Number of Learners"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.3 Unenrollment Duration Analysis

### Duration Statistics

```{r duration-stats, results='asis'}
duration_summary <- all_enrollments %>%
  filter(!is.na(enrollment_duration)) %>%
  summarise(
    Metric = c("Mean", "Median", "Std. Dev", "Min", "Max", "Q1", "Q3"),
    Value_Days = c(
      round(mean(enrollment_duration), 1),
      round(median(enrollment_duration), 1),
      round(sd(enrollment_duration), 1),
      round(min(enrollment_duration), 1),
      round(max(enrollment_duration), 1),
      round(quantile(enrollment_duration, 0.25), 1),
      round(quantile(enrollment_duration, 0.75), 1)
    )
  )

kable(duration_summary,
      caption = "Unenrollment Duration Statistics (Days)",
      align = "lr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Interpretation:** Unenrolled learners persist for an average of **124 days** (median: **72 days**), with high variability (SD: 145 days). This suggests a critical drop-off window around 2–3 months post-enrollment.

### Visualization: Duration Distribution

```{r duration-histogram, fig.cap="Distribution of days from enrollment to unenrollment"}
all_enrollments %>%
  filter(!is.na(enrollment_duration) & enrollment_duration > 0) %>%
  ggplot(aes(x = enrollment_duration)) +
  geom_histogram(binwidth = 30, fill = "#2E86AB", alpha = 0.8, color = "white") +
  geom_vline(aes(xintercept = median(enrollment_duration, na.rm = TRUE)),
             color = "#A23B72", linetype = "dashed", size = 1, label = "Median") +
  geom_vline(aes(xintercept = mean(enrollment_duration, na.rm = TRUE)),
             color = "#F18F01", linetype = "dashed", size = 1, label = "Mean") +
  labs(
    title = "Time to Unenrollment Distribution",
    x = "Days from Enrollment to Unenrollment",
    y = "Number of Learners",
    caption = "Excludes 0-day unenrollments (same-day drop)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  ) +
  scale_x_continuous(breaks = seq(0, 900, 150))
```

### Visualization: Unenrollment by Duration Quartile

```{r duration-quartiles, fig.cap="Breakdown of unenrollments by duration quartiles"}
all_enrollments %>%
  filter(!is.na(enrollment_duration)) %>%
  mutate(
    Duration_Quartile = case_when(
      enrollment_duration <= 7 ~ "≤ 1 week (0-7 days)",
      enrollment_duration <= 30 ~ "1-4 weeks (8-30 days)",
      enrollment_duration <= 72 ~ "1-10 weeks (31-72 days)",
      enrollment_duration <= 180 ~ "2-6 months (73-180 days)",
      TRUE ~ "> 6 months (>180 days)"
    ),
    Duration_Quartile = factor(Duration_Quartile,
                               levels = c("≤ 1 week (0-7 days)",
                                          "1-4 weeks (8-30 days)",
                                          "1-10 weeks (31-72 days)",
                                          "2-6 months (73-180 days)",
                                          "> 6 months (>180 days)"))
  ) %>%
  count(Duration_Quartile) %>%
  mutate(Percentage = round(n / sum(n) * 100, 1)) %>%
  ggplot(aes(x = Duration_Quartile, y = n, fill = Duration_Quartile)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(n, "\n(", Percentage, "%)")),
            vjust = -0.5, size = 3.5) +
  labs(
    title = "Unenrollment by Duration Quartile",
    x = "Time to Unenrollment",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )
```

## 4.4 Engagement by Course Run

### Completion Rates Across Runs

```{r run-comparison, results='asis'}
run_stats <- all_enrollments %>%
  group_by(course_run) %>%
  summarise(
    Enrollments = n(),
    Completed = sum(fully_participated),
    Completion_Rate = round(sum(fully_participated) / n() * 100, 2),
    Unenrolled = sum(unenrolled),
    Unenrollment_Rate = round(sum(unenrolled) / n() * 100, 2),
    Purchased = sum(purchased_statement),
    Purchase_Rate = round(sum(purchased_statement) / n() * 100, 3),
    Avg_Duration = round(mean(enrollment_duration, na.rm = TRUE), 1)
  ) %>%
  rename(
    "Course Run" = course_run,
    "Enroll #" = Enrollments,
    "Complete #" = Completed,
    "Complete %" = Completion_Rate,
    "Unenroll #" = Unenrolled,
    "Unenroll %" = Unenrollment_Rate,
    "Purchase #" = Purchased,
    "Purchase %" = Purchase_Rate,
    "Avg Days" = Avg_Duration
  )

kable(run_stats,
      caption = "Engagement Metrics by Course Run",
      align = "lrrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

### Visualization: Completion Rate by Run

```{r run-completion-plot, fig.cap="Completion rates vary across course runs"}
all_enrollments %>%
  group_by(course_run) %>%
  summarise(
    Completion_Rate = sum(fully_participated) / n() * 100,
    Enrollments = n()
  ) %>%
  ggplot(aes(x = course_run, y = Completion_Rate, fill = course_run)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = paste0(round(Completion_Rate, 2), "%")),
            vjust = -0.5, size = 4, fontface = "bold") +
  labs(
    title = "Course Completion Rate by Run",
    x = "Course Run",
    y = "Completion Rate (%)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(size = 11)
  ) +
  scale_y_continuous(limits = c(0, max(all_enrollments %>%
    group_by(course_run) %>%
    summarise(Completion_Rate = sum(fully_participated) / n() * 100) %>%
    pull(Completion_Rate)) * 1.15))
```


# 5. Key Findings & Insights

## 5.1 Engagement Patterns

1. **Very Low Completion:** Only 5.78% of enrollments result in full course participation, indicating a significant engagement challenge.

2. **Low Monetization:** Certificate/statement purchase rate is extremely low (0.77%), suggesting limited perceived value or willingness to pay among learners.

3. **High Unenrollment:** 11.24% of learners actively unenroll, with a median time of **72 days**—indicating most drop-off occurs in the first 2–3 months.

4. **Large "Unknown" Cohort:** 82.21% of enrollments fall into "Still Active / Unknown," likely representing learners who stop engaging informally without formal unenrollment.

## 5.2 Critical Drop-Off Window

The unenrollment duration analysis reveals:

- **41% of unenrolled learners** drop out within the first week.  
- **Median persistence:** 72 days (approximately 10 weeks).  
- **Max persistence:** 889 days (~2.4 years), indicating some long-term lurkers.  

This suggests an **early intervention window (weeks 1–10)** is critical for retention.

## 5.3 Data Quality Implications

High missing demographic data (>88%) limits demographic segmentation. Analysis cannot confidently compare engagement by age, education, or employment without imputation or assumptions.


# 6. Conclusions & Recommendations

## 6.1 Conclusions

1. **MOOC completion is rare** (~5.8%), and conversion to paid certificates is minimal (~0.8%).  
2. **Most disengagement is informal**—learners don't formally unenroll; they simply stop participating.  
3. **Early weeks are critical**—42% of unenrollments occur within 1 week; the median is ~10 weeks.  
4. **Demographic factors are unmeasured**—missing data prevents deeper segmentation analysis.


This ***Cycle 2*** analysis tests key hypotheses about MOOC learner engagement using statistical methods. Building on Cycle 1's exploratory findings, we employ **one-way ANOVA** to examine whether mean time-to-unenroll (enrollment duration) differs significantly across course runs (H2), chi-squared tests for categorical relationships, and post-hoc tests to identify which groups differ. Key finding: **Mean enrollment duration varies significantly by course run** (p < 0.05), with implications for course-specific retention strategies.


# 7. Cycle 2: Hypothesis Testing Framework

## 7.1 Research Questions & Hypotheses

From Cycle 1, we observed:
- Variable completion rates across 7 course runs (range: 4.8%–7.2%)
- Variable unenrollment durations (median: 72 days)
- Potential demographic differences (though 88% missing)

**Cycle 2 tests three hypotheses:**

| Hypothesis | Question | Test Type | Variables |
|------------|----------|-----------|-----------|
| **H1** | Does mean time-to-unenroll differ by course run? | One-way ANOVA | enrollment_duration ~ course_run |
| **H2** | Does completion rate differ by course run? | Chi-squared test | fully_participated × course_run |
| **H3** | Does mean duration differ by age group (available data)? | One-way ANOVA | enrollment_duration ~ demo_age_group |

**Focus for this report: H1** (mean enrollment duration by course run), with supporting tests for H2 and H3.

## 7.2 CRISP-DM Phase: Evaluation

This cycle addresses **Evaluation** (phase 5 in CRISP-DM):
- Test statistical significance of patterns observed in Cycle 1.
- Determine whether differences are real or due to random variation.
- Prepare conclusions and recommendations for action.

# 8. Data Preparation & Feature Engineering

```{r data-prepare}
# Load raw CSVs
enrollment_1 <- read_csv(file.path(project_root, "data/enrollment_1.csv"))
enrollment_2 <- read_csv(file.path(project_root, "data/enrollment_2.csv"))
enrollment_3 <- read_csv(file.path(project_root, "data/enrollment_3.csv"))
enrollment_4 <- read_csv(file.path(project_root, "data/enrollment_4.csv"))
enrollment_5 <- read_csv(file.path(project_root, "data/enrollment_5.csv"))
enrollment_6 <- read_csv(file.path(project_root, "data/enrollment_6.csv"))
enrollment_7 <- read_csv(file.path(project_root, "data/enrollment_7.csv"))

# Tag course run
enrollment_1 <- enrollment_1 %>% mutate(course_run = 1)
enrollment_2 <- enrollment_2 %>% mutate(course_run = 2)
enrollment_3 <- enrollment_3 %>% mutate(course_run = 3)
enrollment_4 <- enrollment_4 %>% mutate(course_run = 4)
enrollment_5 <- enrollment_5 %>% mutate(course_run = 5)
enrollment_6 <- enrollment_6 %>% mutate(course_run = 6)
enrollment_7 <- enrollment_7 %>% mutate(course_run = 7)

# Combine
all_enrollments <- bind_rows(
  enrollment_1, enrollment_2, enrollment_3, enrollment_4,
  enrollment_5, enrollment_6, enrollment_7
)

# Feature engineering for hypothesis testing
all_enrollments <- all_enrollments %>%
  mutate(
    enrolled_date = as.Date(enrolled_at),
    unenrolled_date = as.Date(unenrolled_at),
    fully_participated = !is.na(fully_participated_at),
    purchased_statement = !is.na(purchased_statement_at),
    unenrolled = !is.na(unenrolled_at),
    enrollment_duration = as.numeric(unenrolled_date - enrolled_date),
    engagement_status = case_when(
      fully_participated ~ "Completed",
      unenrolled ~ "Unenrolled",
      TRUE ~ "Still Active / Unknown"
    ),
    # Demographics (with Unknown handling)
    demo_age_group = case_when(
      age_range %in% c("13-17", "18-24") ~ "Young",
      age_range %in% c("25-34", "35-45", "46-55") ~ "Middle",
      age_range %in% c("56-64", "65+") ~ "Senior",
      TRUE ~ "Unknown"
    ),
    demo_education = case_when(
      highest_education_level %in% c("university_degree", "phd", "masters") ~ "Higher_Ed",
      highest_education_level %in% c("secondary_school", "junior_secondary_school") ~ "Secondary",
      TRUE ~ "Unknown"
    ),
    demo_employment = case_when(
      employment_status %in% c("working_full_time", "working_part_time") ~ "Working",
      employment_status %in% c("unemployed", "not_available") ~ "Not_Working",
      TRUE ~ "Unknown"
    )
  )

cat("Data loaded and prepared.\n")
cat("Total enrollments:", nrow(all_enrollments), "\n")
cat("Course runs:", n_distinct(all_enrollments$course_run), "\n")
```


# 9. Hypothesis H2: Mean Duration Differs by Course Run

## 9.1 Hypothesis Statement

**H1 :**
- **Null Hypothesis (H0):** Mean enrollment duration is equal across all 7 course runs.  
  \(μ_1 = μ_2 = μ_3 = μ_4 = μ_5 = μ_6 = μ_7\)

- **Alternative Hypothesis (H1):** At least one course run has a different mean enrollment duration.  
  \(\text{At least one } μ_i \neq μ_j\)

- **Significance level :** 0.05  
- **Test:** One-way ANOVA

## 9.2 Descriptive Statistics by Course Run

```{r h1-descriptive}
# Calculate duration statistics by course run
duration_by_run <- all_enrollments %>%
  filter(!is.na(enrollment_duration), enrollment_duration >= 0) %>%
  group_by(course_run) %>%
  summarise(
    N = n(),
    Mean_Days = round(mean(enrollment_duration, na.rm = TRUE), 2),
    Median_Days = round(median(enrollment_duration, na.rm = TRUE), 2),
    SD_Days = round(sd(enrollment_duration, na.rm = TRUE), 2),
    Min_Days = min(enrollment_duration, na.rm = TRUE),
    Max_Days = max(enrollment_duration, na.rm = TRUE),
    Q1 = round(quantile(enrollment_duration, 0.25), 2),
    Q3 = round(quantile(enrollment_duration, 0.75), 2),
    .groups = "drop"
  ) %>%
  rename("Course Run" = course_run)

kable(duration_by_run,
      caption = "Enrollment Duration Statistics by Course Run",
      align = "lrrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

**Observation:** Mean enrollment duration ranges from **36 to 180 days** across runs. Run 1 has the highest mean (180.37 days), while Run 7 has the lowest (36.38 days). Visually, differences exist, but are they statistically significant?

## 9.3 Boxplot Visualization

```{r h1-boxplot, fig.cap="Distribution of enrollment duration by course run"}
all_enrollments %>%
  filter(!is.na(enrollment_duration), enrollment_duration >= 0, enrollment_duration <= 500) %>%
  ggplot(aes(x = factor(course_run), y = enrollment_duration, fill = factor(course_run))) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.2, alpha = 0.2, size = 1) +
  labs(
    title = "Enrollment Duration Distribution by Course Run",
    x = "Course Run",
    y = "Days from Enrollment to Unenrollment",
    caption = "Note: Outliers >500 days excluded for clarity"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(size = 11)
  )
```

## 9.4 One-Way ANOVA Test

```{r h1-anova}
# Prepare data for ANOVA (remove NA and extreme outliers)
anova_data <- all_enrollments %>%
  filter(!is.na(enrollment_duration), enrollment_duration >= 0) %>%
  mutate(course_run = factor(course_run))

# Perform one-way ANOVA
anova_result <- aov(enrollment_duration ~ course_run, data = anova_data)
anova_summary <- summary(anova_result)

# Extract key results
f_statistic <- anova_summary[[1]]$`F value`[1]
p_value <- anova_summary[[1]]$`Pr(>F)`[1]
df_between <- anova_summary[[1]]$Df[1]
df_within <- anova_summary[[1]]$Df[2]
ss_between <- anova_summary[[1]]$`Sum Sq`[1]
ss_within <- anova_summary[[1]]$`Sum Sq`[2]
ms_between <- anova_summary[[1]]$`Mean Sq`[1]
ms_within <- anova_summary[[1]]$`Mean Sq`[2]

cat("**One-Way ANOVA Results (H1: Duration by Course Run)**\n\n")
```

### ANOVA Table

```{r h1-anova-table, results='asis'}
anova_table <- tibble(
  Source = c("Course Run", "Error (Within Groups)", "Total"),
  DF = c(df_between, df_within, df_between + df_within),
  `Sum of Squares` = c(ss_between, ss_within, ss_between + ss_within),
  `Mean Square` = c(ms_between, ms_within, NA),
  `F Statistic` = c(f_statistic, NA, NA),
  `p-value` = c(p_value, NA, NA)
)

anova_table <- anova_table %>%
  mutate(
    `Sum of Squares` = round(`Sum of Squares`, 2),
    `Mean Square` = round(`Mean Square`, 2),
    `F Statistic` = round(`F Statistic`, 4),
    `p-value` = round(`p-value`, 6)
  )

kable(anova_table,
      caption = "One-Way ANOVA: Enrollment Duration by Course Run",
      align = "lrrrrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

### ANOVA Interpretation

```{r h1-anova-interpretation}
cat("**Statistical Decision:**\n\n")
cat("- **F-statistic:** ", round(f_statistic, 4), "\n")
cat("- **p-value:** ", format(p_value, scientific = TRUE, digits = 4), "\n")
cat("- **Significance level :** 0.05\n\n")

if (p_value < 0.05) {
  cat("**Conclusion: REJECT the null hypothesis (H0).**\n\n")
  cat("Since p-value (", format(p_value, scientific = TRUE, digits = 3), ") <  (0.05),\n")
  cat("we have **statistically significant evidence** that mean enrollment duration\n")
  cat("**differs across course runs**.\n\n")
  cat("At least one course run has a significantly different mean duration than the others.\n")
} else {
  cat("**Conclusion: FAIL TO REJECT the null hypothesis (H0).**\n\n")
  cat("Since p-value (", format(p_value, scientific = TRUE, digits = 3), ") >=  (0.05),\n")
  cat("we do not have sufficient evidence that mean enrollment duration differs across runs.\n")
}
```
Learners in different runs stay enrolled for different average lengths of time, and this is not just random noise. Some runs (e.g. Run 1) keep learners much longer while others (e.g. Run 7) lose them much sooner.


# 10. Supporting Hypothesis Tests

## 10.1 Hypothesis H2: Completion Rate by Course Run (Chi-Squared Test)

```{r h1-chi-squared}
# Create contingency table
completion_table <- all_enrollments %>%
  group_by(course_run, fully_participated) %>%
  count() %>%
  ungroup() %>%
  mutate(
    course_run = factor(course_run),
    fully_participated = ifelse(fully_participated, "Completed", "Not_Completed")
  ) %>%
  pivot_wider(
    names_from = fully_participated,
    values_from = n,
    values_fill = 0
  )

# Display contingency table
cat("**Contingency Table: Completion Status by Course Run**\n\n")
print(completion_table)

# Perform chi-squared test
chi2_result <- chisq.test(
  x = as.matrix(completion_table[, c("Completed", "Not_Completed")])
)

cat("\n**Chi-Squared Test Results (H1):**\n\n")
cat("- Chi-squared statistic:", round(chi2_result$statistic, 4), "\n")
cat("- p-value:", round(chi2_result$p.value, 4), "\n")
cat("- Degrees of freedom:", chi2_result$parameter, "\n\n")

if (chi2_result$p.value < 0.05) {
  cat("**Conclusion:** Completion rate **significantly differs** across course runs (p < 0.05).\n")
} else {
  cat("**Conclusion:** Completion rate does **not significantly differ** across course runs (p >= 0.05).\n")
}
```

### Completion Rate Table

```{r h2-completion-rates, results='asis'}
completion_rates <- all_enrollments %>%
  group_by(course_run) %>%
  summarise(
    Total = n(),
    Completed = sum(fully_participated),
    Completion_Rate_Pct = round(sum(fully_participated) / n() * 100, 2),
    .groups = "drop"
  ) %>%
  rename("Course Run" = course_run, "Completed #" = Completed, "Rate (%)" = Completion_Rate_Pct)

kable(completion_rates,
      caption = "Completion Rates by Course Run",
      align = "lrrr") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## 10.2 Hypothesis H3: Duration by Age Group (One-Way ANOVA, Available Data)

```{r h3-anova}
# Filter to rows with age group information (not "Unknown")
h3_data <- all_enrollments %>%
  filter(demo_age_group != "Unknown", !is.na(enrollment_duration), enrollment_duration >= 0) %>%
  mutate(demo_age_group = factor(demo_age_group))

if (nrow(h3_data) > 0 && nlevels(h3_data$demo_age_group) >= 2) {
  # Perform ANOVA
  anova_h3 <- aov(enrollment_duration ~ demo_age_group, data = h3_data)
  anova_h3_summary <- summary(anova_h3)
  
  f_h3 <- anova_h3_summary[[1]]$`F value`[1]
  p_h3 <- anova_h3_summary[[1]]$`Pr(>F)`[1]
  
  cat("**One-Way ANOVA: Duration by Age Group (H3)**\n\n")
  cat("- F-statistic:", round(f_h3, 4), "\n")
  cat("- p-value:", round(p_h3, 4), "\n")
  cat("- Sample size (non-Unknown only):", nrow(h3_data), "\n\n")
  
  if (p_h3 < 0.05) {
    cat("**Conclusion:** Mean duration **significantly differs** by age group (p < 0.05).\n")
  } else {
    cat("**Conclusion:** Mean duration does **not significantly differ** by age group (p >= 0.05).\n")
  }
  
  # Descriptive stats
  h3_desc <- h3_data %>%
    group_by(demo_age_group) %>%
    summarise(
      N = n(),
      Mean_Days = round(mean(enrollment_duration), 2),
      SD_Days = round(sd(enrollment_duration), 2),
      Median_Days = round(median(enrollment_duration), 2),
      .groups = "drop"
    ) %>%
    rename("Age Group" = demo_age_group)
  
  cat("\n**Descriptive Statistics:\n\n")
  print(kable(h3_desc, align = "lrrrr", caption = "Enrollment Duration by Age Group"))
} else {
  cat("Insufficient known age group data for H3 analysis.\n")
}
```

# 11. Summary of Hypothesis Tests

```{r hypothesis-summary, results='asis'}
f_h3 <- NA_real_
p_h3 <- NA_real_
hypothesis_summary <- tibble(
  Hypothesis = c("H1", "H2", "H3"),
  Research_Question = c(
    "Mean duration differs by course run?",
    "Completion rate differs by course run?",
    "Mean duration differs by age group?"
  ),
  Test_Used = c("Chi-Squared", "One-Way ANOVA", "One-Way ANOVA"),
  Test_Statistic = c(
    paste0("χ² = ", round(chi2_result$statistic, 2)),
    paste0("F = ", round(f_statistic, 2)),
    paste0("F = ", round(f_h3, 2))
  ),
  P_Value = c(
    format(chi2_result$p.value, scientific = TRUE, digits = 2),
    format(p_value, scientific = TRUE, digits = 2),
    format(p_h3, scientific = TRUE, digits = 2)
  ),
  Decision = c(
    ifelse(chi2_result$p.value < 0.05, "Reject H0 *", "Fail to reject H0"),
    ifelse(p_value < 0.05, "Reject H0 *", "Fail to reject H0"),
    ifelse(p_h3 < 0.05, "Reject H0 *", "Fail to reject H0")
  )
)

kable(hypothesis_summary,
      caption = "Summary of Hypothesis Tests (α = 0.05)",
      align = "llllll") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)

cat("\n*Asterisk indicates statistical significance at α = 0.05 level.")
```

# 12. Key Findings & Interpretation

## 12.1 Main Finding: H2 (Primary Focus)

```{r finding-h1-text}
cat("**H1 RESULT:**\n\n")
cat("Mean enrollment duration **significantly differs** across the 7 course runs.\n\n")

# Identify the runs with highest and lowest mean durations
duration_summary <- all_enrollments %>%
  filter(!is.na(enrollment_duration), enrollment_duration >= 0) %>%
  group_by(course_run) %>%
  summarise(Mean_Days = mean(enrollment_duration), .groups = "drop")

max_run <- duration_summary %>% slice_max(Mean_Days)
min_run <- duration_summary %>% slice_min(Mean_Days)

cat("- **Longest mean duration:** Run ", max_run$course_run, " (~", round(max_run$Mean_Days, 0), " days)\n")
cat("- **Shortest mean duration:** Run ", min_run$course_run, " (~", round(min_run$Mean_Days, 0), " days)\n")
cat("- **Difference:** ~", round(max_run$Mean_Days - min_run$Mean_Days, 0), " days (~4 weeks)\n\n")

cat("**Practical Interpretation:**\n")
cat("Learners in some course runs stay enrolled significantly longer before unenrolling.\n")
cat("This could reflect differences in course pacing, content difficulty, support systems,\n")
cat("or instructor engagement across runs.\n")
```

## 12.2 Supporting Findings: H2 & H3

```{r finding-h2-h3}
cat("**H2 (Completion by Course Run):**\n")
if (chi2_result$p.value < 0.05) {
  cat("Completion rates significantly differ by course run.\n")
  cat("This aligns with H1: runs with longer engagement duration may have higher completion.\n\n")
} else {
  cat("Completion rates do not significantly differ by course run.\n")
  cat("This suggests other factors beyond run may drive completion.\n\n")
}
```

# 13. Recommendations Based on Statistical Findings

## 13.1 Course-Run Specific Actions

```{r recommendations, results='asis'}
recommendations <- tibble(
  Action = c(
    "Review Run 7 design",
    "Extend retention in short-duration runs",
    "Investigate success factors in long-duration runs",
    "Standardize onboarding across runs"
  ),
  Rationale = c(
    "Shortest mean duration (~36 days in Run 7); learners disengage much earlier than in other runs",
    "Runs 5–7 show substantially shorter engagement than Run 1; learners drop out after only a few weeks",
    "Run 1 shows the longest engagement (~180 days); its structure or content may support persistence",
    "Duration ranges roughly from 36 to 180 days; reducing this gap could improve overall consistency"
  ),
  Implementation = c(
    "Audit Run 7 content, pacing, assessments, and support; compare against Run 1’s structure",
    "Introduce milestone emails/check-ins (from weeks 1, 2, 4, 8) in short-duration runs first",
    "Document concrete practices from Run 1 (e.g., scaffolding, feedback frequency) and test them elsewhere",
    "Align key elements (onboarding flow, early activities, communication) with what works best in high-duration runs"
  )
)

kable(recommendations,
      caption = "Data-Driven Recommendations from H2 Results",
      align = "lll") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE)
```

## 13.2 Demographic-Specific Actions (Limited by Data)

```{r demographic-actions}
cat("**Demographic Considerations:**\n\n")
cat("- **Age group:** Limited available data (only 12% of learners provided age).\n")
cat("  Recommendation: Make age an optional question at signup to enable future segmentation.\n\n")
cat("- **Education/Employment:** Similar low response rate (11–12% complete demographic data).\n")
cat("  Recommendation: Incentivize demographic completion (e.g., certificate milestone) in Cycle 3.\n\n")
cat("Action: Collect richer demographics in next cycle to enable targeted support by age/background.\n")
```

# 14. CRISP-DM Cycle 2 Evaluation Summary


## 14.1 Key Statistical Outputs

| Test | Statistic | p-value | Conclusion |
|------|-----------|---------|------------|
| **H2: ANOVA (Duration × Course Run)** | F = `r round(f_statistic, 4)` | `r format(p_value, scientific=TRUE, digits=2)` | **Significant*** |
| **H1: Chi-Squared (Completion × Course Run)** | χ² = `r round(chi2_result$statistic, 2)` | `r format(chi2_result$p.value, scientific=TRUE, digits=2)` | **Significant*** |
| **H3: ANOVA (Duration × Age Group)** | F = `r round(f_h3, 4)` | `r format(p_h3, scientific=TRUE, digits=2)` | Varies |

*p < 0.05; results are statistically significant at the 5% level.

# 15. Conclusion

## 15.1 Summary

This Cycle 2 analysis confirms that **mean enrollment duration is statistically significantly different across course runs** (H1, p < 0.05). Learners in Run 1 disengage ~27 days earlier than learners in Run 5, a difference that is both statistically and practically meaningful.

Supporting evidence shows:
- **H2 (Completion by Run):** Also significant, suggesting run quality directly impacts engagement outcomes.  
- **H3 (Age Group):** Limited data, but no strong effect detected among available respondents.  

## 15.2 Limitations

1. **Missing Demographics:** 88% of learners did not provide age/education/employment data, limiting segmentation analysis.  
2. **No Causal Mechanism:** Statistical significance shows *that* differences exist, not *why* they exist.  
3. **No Experimental Control:** This is observational; cannot claim Run 1 changes will raise duration to Run 5 levels without intervention testing.